---
title: "Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=6) 
library(isoSIM)
```

# ISOSIM
The package isoSIM was designed to simulate the generation of iso-female lines, and simulate genomic changes during, and after, formation of iso-female lines. Furthermore, it allows for simulation of effects after crossing individuals from iso-female lines.

## Simulating an iso-female line
In order to create an iso-female line, we first have to create a 'wild' population, from which we will draw individuals. To create such a population, we make use of the function 'create_population':
```{r}
wildpop =  create_full_population(pop_size = 100, 
                             number_of_founders = 10,
                             total_runtime = 1000, 
                             morgan = 1, 
                             seed = 42, 
                             write_to_file = FALSE)

```
This creates a population of 100 diploid individuals (e.g. 2N = 200), based upon 10 founders (e.g. 10 individuals that can be genetically distinguished from each other). Then, for 1000 generations, this population is inbred and genomes of the 20 founders are allowed to mix. All individuals have 2 chromosomes (for simplicity and computational speed, we only model one chromosome pair), which are 1 Morgan long. The result can be written to file (optional), which can be usefull when generating for instance an extremely large wild population, which can be used later (e.g. read from file).

The result of the function is a structure containing 1000 individuals:
```{r}
wildpop
```
Each individual has two chromosomes, with a given number of junctions (e.g. delineations between two contiguous genomic stretches from different ancestors):
```{r}
wildpop[[1]]
```

In order to create an iso-female line, two random individuals are drawn from the wild population, and selected for inbreeding. We can simulate this process with:
```{r}
isofemale <- create_iso_female(source_pop = wildpop,
                               n = 1,
                               inbreeding_pop_size = 100,
                               run_time = 1000000,
                               morgan = 1)
```
Here, n indicates the number of isofemales to be created from the same source population, the inbreeding population size is set to be small, in order to speed up computation. Maximum run time is best set hight, to assure that all individuals become genetically identical. 

# Visualizing individuals
Now, we can plot the isofemales, this plots the two chromosomes next to each other, where colors indicate different ancestors:
```{r}
plot(isofemale[[1]])
```

Because we have chosen 10 ancestors, these plots are not terribly informative, let's try a toy example:
```{r}
wildpop =  create_full_population(pop_size = 100, 
                             number_of_founders = 2,
                             total_runtime = 10, 
                             morgan = 1, 
                             seed = 666, 
                             write_to_file = FALSE)

isofemale <- create_iso_female(source_pop = wildpop,
                               n = 1,
                               inbreeding_pop_size = 100,
                               run_time = 1000000,
                               morgan = 1)
plot(wildpop[[1]])
plot(isofemale[[1]])
```
We can also plot a (section of) a single chromosome:
```{r}
plot_chromosome(isofemale[[1]]$chromosome1, xmin = 0.0, xmax = 0.5)
```













